Convergence rate of MAP estimates \\ for the exponential family

### 2021-10-15 aistats.tex
Pass on introduction.

{\bf Exponential families} are among the most widely used statistical models.
Many standard random variables fall into this category: Gaussians, categorical, Gamma or Dirichlet.
They are flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
In particular, they are appreciated for their convex log-likelihood
\alignn{
f(\nat) = \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
\label{eq:defNLL}
}
where $\logpart$ is a convex function and $\nat\in\Theta$ is a vector called \textit{natural parameter}.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized}
or variants of principal component analysis \citep{collins2001generalization}, among other applications.

{\bf Estimators.}
In general, we want to estimate $\nat$ from some data $\mD = (x_1, \dots, x_n)$.
For this purpose, not only is $f$ convex, but it even enjoys a closed form maximum-likelihood estimate (MLE)
\begin{align}
	\nabla \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
	\label{eq:defMLE}
\end{align}
This rule is known as moment matching.
Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP).
The parameter $\theta$ can also be estimated with online-learning techniques, but they are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.

{\bf Metrics.}
Once we have an estimator $\hat \nat$, we wish to assess its quality, i.e., its closeness to some optimal $\nat^*$.
This is necessary, for instance to choose between 2 estimators (model selection).
We distinguish here two ways: \textit{disntance in parameter space} and \textit{distance between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \&p(x)>0$.
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
Hence the fundamental problem:
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

We have already asymptotic solutions (\S\ref{ssec:asymptote}) when $\logpart$ is quadratic (e.g. $X$ is Gaussian with known variance)
or close to quadratic (\S\ref{ssec:quadratic}).
However, a  general solution for finite $n$ remains elusive.
In this paper, we review partial solutions, and give ideas on how to solve it.

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)\,,
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} }
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is an instantiation of stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we try using existing optimization convergence rates for SMD to get the upper bound we seek.
Unfortunately, none of the current analyses apply, which frames another open problem for the analysis of SMD.

{\bf Expected Ouctomes.}
A solution to~\eqref{problem} would quantify the importance of the prior in MAP, in particular in the few sample regime.%, for instance for Gaussians $\cN(m,\sigma^2)$.
Also, it could enable stochastic optimization to tackle a wide class of barrier objectives\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}.
A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
It could also help in assessing the impact of alternative forms of regularization (prior) for these models.

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}) .
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
	\item By linking MAP and SMD, we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
\end{itemize}

>>>>

{\bf Exponential families} are among the most widely used simple parametric models of data, yet, we will highlight some open problems about them in this paper.
Many standard random variables are exponential families: Gaussians, categorical, gamma or Dirichlet for example.
They are flexible enough to model a variety of data sources $X$, and easy to describe with some sufficient statistics $T(X) \in \real^d$.
They are particularly appreciated for their convex log-likelihood
\alignn{
f(\nat) := \E[-\log p_\nat(X)] = \logpart(\nat) - \lin{\E[T(X)] , \nat},
\label{eq:defNLL}
}
where $\logpart$ is the convex log-partition function and $\nat\in\Theta$ is the \textit{natural parameter}.
This convexity lays the foundation for generalized linear models \citep{mccullagh1989generalized}
or variants of principal component analysis \citep{collins2001generalization}, among other applications.

{\bf Estimators.}
In this paper, we consider the problem of estimating $\nat$ from a dataset $\mD = (X_1, \dots, X_n)$ of iid observations from $p_\theta$ in an exponential family.
In this case, not only is $f$ convex, but it yields a simple condition for the maximum-likelihood estimate (MLE)
\begin{align}
 \hat \mu_n^\text{MLE} = \nabla  \logpart(\hat \nat_n^\text{MLE}) = \frac{\sum_{i=1}^n T(x_i)}{n} \; .
	\label{eq:defMLE}
\end{align}
This rule is also known as moment matching.
Given a certain conjugate prior, a similar formula~\eqref{eq:defMAP} holds for the maximum a posteriori (MAP). In this paper, we will focus on analyzing MLE and MAP estimators.\footnote{A related analysis is present in the online-learning literature, but for different online estimators which are less efficient than offline methods \citep{azoury2001relative,dasgupta2007online}.}

{\bf Statistical decision theory.}
To assess the quality of an estimator $\hat \nat$ (and compare them), we need to define some notion of closeness to the correct parameter $\nat^*$.
We distinguish here two ways: \textit{distance in parameter space} and \textit{``distance'' between distributions}.
{\bf 1)} Distance in parameter space $d(\nat,\nat^*)$. This is the focus of \emph{parameter estimation}, yielding results such as the asymptotic efficiency of the MLE via the Cramer-Rao lower-bound \citep{aitken1942estimation} and a wealth of asymptotic results \citep{vdv1998asymptotic}.
In particular for sum of independent variables such as~\eqref{eq:defMLE}, large deviations theory \citep{varadhan1984large} characterizes concentration phenomena.
{\bf 2)} Distance between distributions, as studied in \emph{density estimation}.
For this purpose, the Kullback-Leibler (KL) divergence $\KL\paren{p_{\nat^*} || p_{\nat} }$  arises naturally from information theory,
but its lack of robustness to misspecification\footnote{
For $p$ and $q$ continuous densities,
$\KL(p||q) = +\infty$ if $\exists x, q(x)=0 \, \& \, p(x)>0$.
%Since exponential families have a positive density over their whole input set, this is not a concern for us, except if we misspecified the input set.
}
has led statisticians to study symmetric, better-behaved distances, such as the $L^2$ norm \citep[\S1.2]{tsybakov2009introduction}, the $L^1$ norm \citep{devroye2001combinatorial} or more recently the Hellinger distance \citep{baraud2017new}.
With exponential families, the KL divergence is also a Bregman divergence between parameters (see \S\ref{sec:problem}), thus drawing a connection between these two lines of research.
We thus consider the fundamental problem:
%\begin{center}
%	\framebox{
%		\parbox[c][]{0.9\linewidth}{
%			\begin{center}
%			\emph{Can we upper bound  the expected value of the KL divergence for MLE or MAP?}
%		\end{center}
%		}
%	}
%\end{center}
\begin{equation}
\boxed{
\begin{aligned}
	\textit{Find an upper}&\textit{ bound on the expected value of } \\
	&\KL(p_{\nat^*} || p_{\hat \nat_n^\text{MLE/MAP} }) \; .
\end{aligned}
}
\tag{$\star$}
\label{problem}
\end{equation}
% RLP : that is a dirty hack, but I would like to be able to reference the problem somehow.

There are already general asymptotic results (\S\ref{ssec:asymptote}), and a finite $n$ result when $\logpart$ is quadratic (e.g. $X$ is Gaussian with known variance)
or close to quadratic (\S\ref{ssec:quadratic}).
However, a  general solution for finite $n$ remains elusive.
In this paper, we review partial solutions, and give ideas on how to solve it.

{\bf Stochastic optimization} offers an interesting perspective on this question.
Consider the problem
\alignn{
	\min_{\theta\in \Theta} f(\theta)\,,
	\label{eq:optimization_problem}
}
solved by $\nat^*\in \Theta$.
Setting $f$ as the log-likelihood~\eqref{eq:defNLL}, the suboptimality is equal to the KL:
\alignn{
	f(\nat) - f(\nat^*) = \KL\paren{p_{\nat^*} || p_{\nat} } .
	\label{eq:suboptimalityKL}
}
Both MLE and MAP can be seen as stochastic algorithms solving~\eqref{eq:optimization_problem}.
In particular, with exponential families, MAP is an instantiation of the so-called stochastic mirror descent (SMD) \citep{nemirovski2009robust}.
Inspired by recent work~\citep{lepriol2021analysis, kunstner2020homeomorphic}, we consider using existing optimization convergence rates for SMD to get the upper bound we seek.
Unfortunately, none of the current analyses apply, highlighting open problems for the analysis of SMD.

{\bf Expected Outcomes.}
A solution to~\eqref{problem} can clarify the importance of the prior in MAP, in particular in the few sample regime. %, for instance for Gaussians $\cN(m,\sigma^2)$.
Also, it could enable stochastic optimization to tackle a wide class of barrier objectives.\footnote{we call \emph{barrier} an objective $f$ that is infinite on the boundaries of its domain (assuming they exist).}
A good example would be generalized linear model based on Gaussians with unknown mean and variance, for which there is currently no theory \citep{bach2013nonstronglyconvex}.
It could also help in assessing the impact of alternative forms of regularization (prior) for these models.

{\bf Contributions.}
After formalizing the problem~\eqref{problem} (\S\ref{sec:problem}), along with its asymptotic properties (\S\ref{ssec:asymptote}), we make the following contributions.
\begin{itemize}
	\itemsep0em
	\item We provide an upper bound on the KL in the special case of a Gaussian with known mean but unknown variance $\cN(0,\sigma^2)$ (\S\ref{ssec:gaussian-variance}), illustrating that tight rates are possible even though the current theory does not cover them.
	\item We highlight sufficient conditions to characterize when a (local) quadratic approximation of the KL is valid, offering a partial answer to~\eqref{problem} (\S\ref{ssec:quadratic}-\ref{ssec:local-quadratic}).
	%SLJ REMOVED per Slack discussion
	%\item We leverage a generalized bias-variance decomposition of Bregman divergences\citep{pfau2013generalized}  to formulate a conjecture about~\eqref{problem}(\S\ref{ssec:bias-variance}).
	\item By linking MAP and SMD, we show modern analysis of SMD are yet to prove convergence on barrier objectives such as $-\log$ (\S\ref{sec:optimization}).
\end{itemize}
